Visual Odometry for Autonomous Underwater Vehicles

### 2013-04-08 paper.tex
some minor fixes

Previous to the motion estimation stage, keypoint bucketing \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, the motion estimation process can be solved through different methods. Using a closed-form solution to the least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008,Geiger2011,Huan2011}.

@@

Figure \ref{sample-images} shows some example images from the laboratory and harbour enviroments. Ground truth for the laboratory experiment is extracted by matching each image that has been captured against the known image that is printed on the floor of the test pool. As the size of the print is known, 6 DOF camera poses can be computed minimizing reprojection errors of matched features.

@@

In the harbour experiment, a DVL and an AHS (Airborne Hyperspectral Scanner) have been used as navigation sensors to let the vehicle follow a previously defined trajectory autonomously. In the laboratory experiments, the vehicle has been remotely controlled by a human operator. Table \ref{experiment-details} summarizes the characteristics of the experiments.

@@

As visual odometry suffers from drift, comparing whole trajectories, i.e. poses in time directly to ground truth is not very meaningful. Instead, we subdivide the paths into small pieces and compare velocities for each piece to the matching piece in our ground truth. This method is widely used in visual odometry evaluation \cite{Kitti2013} and depends clearly on the size of the subpaths. However, since it is used to compare different visual odometers, the important thing here is to be coherent with the size of the pieces for all algorithms.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the bottom ones refer to harbour tests. Linear velocity is shown on the left for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

@@

The results presented in Table \ref{results-errors} demostrates that libviso2 has similar mean translation error for both enviroments, but differs significantly in rotation error. By refering to Table \ref{results-runtime}, the mean number of inliers per frame is larger in the laboratory than in the harbour (where environmental conditions are worse). It means libviso2 presents a strong performance in the linear motion estimation when the number of inliers is reduced.

Results for fovis odometer are substantially different in translation error comparing laboratory and harbour conditions. By examining the number of inliers of the two odometers shown in Table \ref{results-runtime} it is easy to see that fovis is significantly less robust in harbour enviroments where the quality of the images is lower and the seabed has a poor texture (see Figure \ref{sample-images}). Furthermore, unlike libviso2, fovis presents a better performance when estimating the angular motion in harbour conditions. This improvement over libviso2 is due to the method of keyframe selection based on the number of matches introduced by fovis. This technique has a positive impact when the vehicle experiences pure rotational movements, as it can keep the reference frame for longer and therefore reduce the drift significally.

@@

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of translational and rotational errors for laboratory and harbour experiments.}
  \label{results-errors}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Enviroment} & \textbf{Algorithm} & \textbf{Translation} & \textbf{Rotation}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 0.978\% & 0.001739 [deg/m] \\
                                & fovis    & 0.601\% & 0.006311 [deg/m] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 0.833\% & 0.007913 [deg/m] \\
                                & fovis    & 1.049\% & 0.003122 [deg/m] \\ \hline
  \end{tabular} 
\end{table}

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of number of inliear and runtime for laboratory and harbour experiments.}
  \label{results-runtime}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Enviroment} & \textbf{Algorithm} & \textbf{Num. inliers} & \textbf{Runtime}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 366 & 0.106 [s] \\
                                & fovis    & 500 & 0.246 [s] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 137 & 0.086 [s] \\
                                & fovis    & 64 & 0.058 [s] \\ \hline
  \end{tabular} 
\end{table}

@@

Despite the mentioned drawbacks we have shown that a visual odometer using a downward looking stereo camera can be a valuable sensor for underwater vehicles giving good estimates for linear movement and rotation about the vehicle's vertical axis. Comparing the performance of libviso2 and fovis in underwater enviroments, libviso2 is preferible when the vehicle operates in real marine waters with critical lighting conditions and slightly textured seabed. The robustness of this algorithm against poor visibility contrasts with the fact of having a worse rotation estimation when compared with fovis in harbour experiments.

A possible improvement for the angular motion estimation in libviso2 could be the implementation of the reference frame selection technique developed by fovis algorithm. This method could substancialy reduce the drift in both linear and angular estimates.

@@

This work is partially supported by the Spanish Ministry of Economy and Competitiveness under contract DPI2011-27977-C03- 02 (TRITON Project), FEDER Fundings and by the European Commissions FP7 under grant agreement 248497 (TRIDENT Project).

>>>>

Previous to the motion estimation stage, keypoint bucketing \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, the motion estimation process can be solved through different methods. Using a closed-form solution to the least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008,Geiger2011,Huang2011}.

@@

Figure \ref{sample-images} shows some example images from the laboratory and harbour environments. Ground truth for the laboratory experiment is extracted by matching each image that has been captured against the known image that is printed on the floor of the test pool. As the size of the print is known, 6 DOF camera poses can be computed minimizing reprojection errors of matched features.

@@

In the harbour experiment, a DVL and an AHRS have been used as navigation sensors to let the vehicle follow a previously defined trajectory autonomously. In the laboratory experiments, the vehicle has been remotely controlled by a human operator. Table \ref{experiment-details} summarizes the characteristics of the experiments.

@@

As visual odometry suffers from drift, comparing whole trajectories, i.e. poses in time directly to ground truth is not very meaningful. Instead, we subdivide the paths into small pieces and compare velocities for each piece to the matching piece in our ground truth. This method is widely used in visual odometry evaluation \cite{Kitti2013} and depends clearly on the size of the sub-paths. However, since it is used to compare different visual odometers, the important thing here is to be coherent with the size of the pieces for all algorithms.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the bottom ones refer to harbour tests. Linear velocity is shown on the left for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and ground truth (green) is also presented. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory environment.

@@

The results presented in Table \ref{results-errors} demonstrates that libviso2 has similar mean translation error for both environments, but differs significantly in rotation error. As can be seen in Table \ref{results-runtime}, the mean number of inliers per frame is larger in the laboratory than in the harbour (where environmental conditions are worse). It means libviso2 presents a strong performance in the linear motion estimation even when the number of inliers is small.

Results for the fovis odometer are substantially different in translation error comparing laboratory and harbour conditions. By examining the number of inliers of the two odometers shown in Table \ref{results-runtime} it is easy to see that fovis is significantly less robust in harbour environments where the quality of the images is lower and the seabed has a poor texture (see Figure \ref{sample-images}). Furthermore, unlike libviso2, fovis presents a better performance when estimating the angular motion in harbour conditions. This improvement over libviso2 is due to the method of keyframe selection based on the number of matches introduced by fovis. This technique has a positive impact when the vehicle experiences pure rotational movements, as it can keep the reference frame for longer and therefore reduce the drift significantly.

@@

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of translational and rotational errors for laboratory and harbour experiments.}
  \label{results-errors}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Environment} & \textbf{Algorithm} & \textbf{Translation} & \textbf{Rotation}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 0.978\% & 0.001739 [deg/m] \\
                                & fovis    & 0.601\% & 0.006311 [deg/m] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 0.833\% & 0.007913 [deg/m] \\
                                & fovis    & 1.049\% & 0.003122 [deg/m] \\ \hline
  \end{tabular} 
\end{table}

\begin{table}[!t]
  \renewcommand{\arraystretch}{1.3}
  \caption{Comparison of number of inliers and runtime for laboratory and harbour experiments.}
  \label{results-runtime}
  \centering
  \begin{tabular}{|l|l|c|c|} \hline
    \textbf{Environment} & \textbf{Algorithm} & \textbf{Num. inliers} & \textbf{Runtime}\\ \hline
    \multirow{2}{*}{Laboratory} & libviso2 & 366 & 0.106 [s] \\
                                & fovis    & 500 & 0.246 [s] \\ \hline
    \multirow{2}{*}{Harbour}    & libviso2 & 137 & 0.086 [s] \\
                                & fovis    & 64 & 0.058 [s] \\ \hline
  \end{tabular} 
\end{table}

@@

Despite the mentioned drawbacks we have shown that a visual odometer using a downward looking stereo camera can be a valuable sensor for underwater vehicles giving good estimates for linear movement and rotation about the vehicle's vertical axis. Comparing the performance of libviso2 and fovis in underwater environments, libviso2 is preferable when the vehicle operates in real marine waters with critical lighting conditions and slightly textured seabed. The robustness of this algorithm against poor visibility contrasts with the fact of having a worse rotation estimation when compared with fovis in harbour experiments.

A possible improvement for the angular motion estimation in libviso2 could be the implementation of the reference frame selection technique developed by fovis algorithm. This method could substantially reduce the drift in both linear and angular estimates.

@@

This work is partially supported by the Spanish Ministry of Economy and Competitiveness under contract DPI2011-27977-C03-02 (TRITON Project), FEDER Fundings and by the European Commissions FP7 under grant agreement 248497 (TRIDENT Project).
